### 2023-02-06 ESS_Trzesicki/Trzesicki.tex
Final commit before sending to the editors for corrections

\documentclass[pdftex,12pt]{article}
% PP \usepackage{apacite}
\usepackage[utf8]{inputenc}
\usepackage[russian,greek,polish,english]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsxtra}
\usepackage{amstext}
%\usepackage{multind} %trzeba przywrócić, kiedy rob się index %PP

%PP \usepackage{url}
\usepackage{graphics,epsf}
\usepackage{aurical}
\usepackage{mathpazo}
\usepackage{wrapfig} %\begin{wrapfigure}[lineheight]{alignment}{width}
\usepackage[pdftex]{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{pstricks}
\usepackage{tabularx}
\usepackage{endnotes}

\usepackage{multicol}
\usepackage{setspace}
\usepackage{pst-plot}
\usepackage{pstricks-add}
\usepackage{pst-func,pstricks-add}
\usepackage{xcolor}
\usepackage{pgfplots}

\usepackage[utopia]{mathdesign}
\usetikzlibrary {plothandlers}
\usetikzlibrary{plotmarks}

\usetikzlibrary{decorations}

\frenchspacing

\usepackage{icomma}

%begin of PP addition ---------------
%\usepackage{natbib}
\usepackage{hyperref}

\usepackage[style=bath,
citestyle=authoryear-comp,
backend=biber]{biblatex}
\addbibresource{Trzesicki.bib}
%end ----------------

% PP \bibliographystyle{apacite}

\usepackage{mathpazo}
\usepackage{lmodern} %dopisane

%\newcommand{\kontr}{\begin{sideways}{\begin{sideways}$\models$\end{sideways}}\end{sideways}}

\usepackage{rotating}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{authblk}

%\renewcommand{\notesname}{Przypisy}

\font\grmn = grmn1000 at 12pt

\usepackage{chngcntr}

%\let\footnote=\endnote %PP

\widowpenalty=300
\clubpenalty=300

\pgfplotsset{compat=1.16}

\setcounter{tocdepth}{6}

%%%%%%%%%%%%%%%

%\usepackage{csquotes}% Recommended

%\usepackage[style=authoryear-ibid,backend=biber]{biblatex}

%%%%%%%%%%%%%%%%%

\begin{document}
\tolerance=1000

%PP {\titlepage
\begin{center}
Kazimierz Trzęsicki

\medskip
\medskip
\textbf{Perspective on Turing Paradigm}
\end{center}
\bigskip
%PP }

%PP \maketitle

%PP \setcounter{tocdepth}{6}
%PP \tableofcontents 
%PP \medskip

%PP \newpage
\hfill \begin{minipage} {10cm} \small

\textit{Tolle numerum omnibus rebus, et omnia pereunt.} \\ (Take from all things their number and all shall perish.)

\smallskip

  St. Isidore of Seville, Patron saint for the Internet

  \emph{Etymologiarum sive originum} \parencite*[Liber III, \textit{De mathematica}, IV. \textit{Quid praestent numeri}]{Izydor1911}

 \end{minipage} \\

\bigskip

\begin{abstract}

Scientific knowledge is acquired according to some paradigm. Galileo wrote that the Book of Nature was written in mathematical language and could not be understood unless one first understood the language and recognized the characters with which it was written. It is argued that Turing planted the seeds of a new paradigm. According to the Turing Paradigm, the Book of Nature is written in algorithmic language, and science aims to learn how the algorithms change the physical, social, and human universe. Some sources of the Turing Paradigm are pointed out, and a few examples of the application of the Turing Paradigm are discussed.

\textbf{Keywords}: Galileo Galilei, Alan Turing, Konrad Zuse, zero, Arabic numeral, paradigm, mathematics, algorithmics
\end{abstract}

@@

Science is developed and created  according to a pattern, a paradigm. Paradigm is historically mutable. The place of one paradigm is taken by the one which enables a fuller understanding and a better description of the information obtained. The Aristotle paradigm of natural knowledge has been replaced by the paradigm  we used to tie with Galileo. According to this paradigm modern science was created. Appointed by Aristotele paradigm of logic lasted until Gottlob Frege. The Greek paradigm of mathematics has been replaced by a paradigm that we can associate with Descartes.

@@

It is usually stated that binary notation was invented and first formally proposed by Leibniz as an illustration of his dualistic philosophy, but already around 1600 the notation was used by  an English astronomer Thomas Harriot. John Shirley \parencite*{Shirley1951} writes about his achievements:  \begin{quote} \small the mathematical papers of Thomas Harriot (1560--1621) show clearly that Harriot not only experimented with number systems, but also understood clearly the theory and practice of binary numeration nearly a century before Leibniz's time.  \end{quote} 
Several manuscripts of the legacy of Thomas Harriot are evidence that he is probably the first inventor of binary system. He uses $0$ and $1$ and shows examples how to convert expressions  written in the decimal system to expressions written in the binary system and \emph{vice versa}. He demonstrates the basic arithmetic operations, too \parencite{Ineichen2008}. As the first text on the binary system Ineichen  points the two-volume work \emph{Mathesis biceps vetus et nova} \parencite*{Lobkowitz1670} by Juan Caramuel y Lobkowitz (Ioannis Caramuelis). In connection with these works by Harriot and Caramuel, the question is raised as to whether Leibniz plagiarized. This question is answered in the affirmative \parencite{Ares2018}.

@@

The use of the binary system in computers was finally determined by \emph{Burks-Goldstine-Von Neuman Report} to U.S. Army Ordnance Department (finished June 28, 1946) reprinted in \parencite*[p.105]{BurksGoldstineNeumann1947}, in which we read: \begin{quote} \small An additional point that deserves emphasis is this: An important part of the machine is not arithmetical, but logical in nature. Now logics, being a yes-no system, is fundamentally binary. Therefore, a binary arrangement of the arithmetical organs contributes very significantly towards a more homogeneous machine, which can be better integrated and is more efficient.\footnote{NB. This explanatory passage was not present in first edition of the report \parencite[cf.][p.13]{burks_preliminary_1946}, but was added in later editions (editor's footnote).}
\end{quote}

@@

The idea that everything is made of $0$ and $1$ is the reason why one of the creators of algorithmic information theory, Gregory Chaitin---as he writes not quite seriously---proposes to name the basic unit of information not ``bit'' but ``leibniz'' \parencites{Chaitin2004}[cf.][]{Trzesicki2006a}: \begin{quote} \small all of information theory derives from Leibniz, for he was the first to emphasize the creative combinatorial potential of the $0$ and $1$ bit, and how everything can be built up from this one elemental choice, from these two elemental possibilities. So, perhaps not entirely seriously, I should propose changing the name of the unit of information from the bit to the leibniz! \end{quote} The unit ``leibniz'' could be the unit (parcel) that Hobbes \parencite*[Chapter V. Of Reason, and Science]{Hobbes1651} wrote about: \begin{quote} \small When a man reasoneth, hee does nothing els but conceive a summe totall, from Addition of parcels. \end{quote}

Leibniz was convinced that the world was organized according to the rules of mathematics. This thought is summarized in the sentence \parencite*[p.191] {Leibniz1677}\footnote{More on this entry in the margin of the essay \emph{Dialogus} \parencite{Leibniz1677} see \parencite{Kopania2018}.}: \begin{quote} \small \emph{Cum Deus calculat et cogitationem exercet, fit mundus}. \end{quote} Mathematics is a tool of the World Constructor, and numbers are the material the world is made of.

@@

The shaping of the modern paradigm of science in what was then called ``natural philosophy'' was in fact a revival of the concept of Archimedes \parencite[pp.71, 77] {Heller2013}. This idea continued in the Middle Ages. For Roger Bacon (1214--1292) there are four great sciences without which others cannot be known and the meaning of things cannot be understood. And when they are known, then wisdom will be attained without difficulty and labor, not only in the teachings of man, but also in the divine ones. And the possibilities of each of these sciences  are revealed not only because of the wisdom itself, but also in relation to the above-mentioned one. In  \emph{Opus Majus} \parencite*[Pars Quarta, Distinctio Prima, Capitulum I]{Bacon2010a} Roger Bacon emphasizes that: \begin{quote} \small Of these sciences the gate and key is mathematics, which the saints discovered at the beginning of the world, as I shall show, and which has always been used by all the saints and sages more than all other sciences. Neglect of mathematics works injury to all knowledge, since he who is ignorant of it cannot know the other sciences or the things of this world. And what is worse, men who are thus ignorant are unable to perceive their own ignorance and so do not seek a remedy. \end{quote} About the place and role of experiments in \emph{De scientia experimentorum: que dicitur dignior Omnibus Partibus Philosophie Naturalis de Perspective: Et ideo notanda est maxime}, a part of \emph{Opus Tertium} \parencite*{Bacon1912}, he wrote that the strongest argument proves nothing so long as the conclusions are not verified by experience. Experimental science is the queen of sciences, and the goal of all speculation.

Galileo justifies heliocentrism by referring to the exegesis of \emph {Bible} based on the doctrine of St. Augustine, in particular his \emph{De Genesi ad litteram}
%PP \footnote{\url{http://www.inters.org/galilei-madame-christina-Lorraine} [02.04.2019]} 
\parencites{galileo_galilei_letter_1615}[cf.][p.73]{Sibley2013}. In this tradition, which Galileo finds explicitly, the book of nature should be read with mathematical tools rather than those of scholastic philosophy. The Book of Nature was written in the language of mathematics, and therefore must be interpreted by mathematicians, not theologians. The book of nature as a mathematics contains truths that cannot be discussed.

Galileo \parencite*[p.4]{Galileo1960} writes: \begin{quote} \small Philosophy is written in this grand book, the universe, which stands continually open to our gaze. But the book cannot be understood unless one first learns to comprehend the language and read the letters in which it is composed. It is written in the language of mathematics, and its characters are triangles, circles, and other geometric figures without which it is humanly impossible to understand a single word of it; without these, one wanders
about in a dark labyrinth. \end{quote} Galileo recommends learning the language of mathematics because it is the language spoken by God \parencite{Wouk2010,Strogatz2019}.

Hall \parencite*[p.97]{Hall1956}  maintains that: \begin{quote} \small  Galileo's greatest fame is as an astronomer, yet in intellectual quality and weight his one treatise on mechanics almost outweighs all the rest of his writings. Although his book on cosmology became notorious, and had a more general public influence, it had no comparable effect upon the future development of scientific astronomy, for its polemics were suited only to its own age. The contradiction here is more apparent than genuine. Though formally divided between two branches, Galileo's creative activity in science was a unity, not twofold: it was a unity in laying down revised principles of procedure in science, and again in its specific exemplification of these principles, since Galileo saw that the science of motion and the just appraisal of the results of observational astronomy were the twin keys to an understanding of the universe. \end{quote}

@@

Newton creates calculus because it is the language with which the book of nature is written. Also Leibniz creates calculus. As an aside, let us add that Newton accused him of plagiarism \parencite{Sonar2018}. It just so happens that Leibniz, the genius of creating symbols \parencite['The Symbol Master', cf.][pp.165--168]{Mazur2014}---having a greater understanding of the choice of language---gave his version a linguistic representation that resulted in the development of which failed with the approach proposed by Newton. Charles Babbage, the creator of the first (mechanical) programmable computer, noticing the delay of English mathematics in relation to French mathematics, undertook to translate French texts from mathematics \parencite{Trzesicki2006c}. He wrote \parencite*{Babbage1864,Babbage2008}: \begin{quote} \small Under these circumstances it was not surprising that I should perceive and be penetrated with the superior power of the notation of Leibniz.  \end{quote}

@@

Later the idea of God (God of Spinoza) as a ``mathematician'' is proclaimed by Einstein \parencite[p.279]{Infeld1980}: \begin{quote} \small God does not care about our mathematical difficulties. He integrates empirically. \end{quote} This is according to Heller \parencite*[p.41]{Heller2014}: \begin{quote} \small A fundamental hypothesis, tacitly taken in the very method of modern mathematized empirical science [which] states that there is nothing in the material world that cannot be mathematically described. \end{quote}

@@

Robert of Chester, who was the first translator into Latin of the now-lost book al-Khwarizmi \parencite[p.411]{Menninger1969}, his translation---found in the 19\textsuperscript{th} century---begins with the words: \begin{quote} \small \emph {Dixit Algoritmi: laudes deo rectori nostro atque defensori dicmus dignas.} \end{quote} Around 1143 \parencite[p.411]{Menninger1958} an abstract was made of this work today known as \emph{Salem Codex}  \parencite{Cantor1865}. At the beginning we read: \begin{quote} \small \emph {Incipit liber algorizmi: omnis sapientia sive scientia a domine Deo; sicut scriptum est: Hoc quod continent omnia scientiam habet, et iterum: Omnia in mensura et pondere et numero constituisti.} \end{quote} The grammatical form used proves that the author was not aware that it was a surname \parencite[p.14, footnote 1]{Cantor1865}. In this text, the name ``algorizmi'' was for the first time---in the preserved literature---used to mark a procedure: \begin{quote} \small Der Gebrauch des Nominativus \emph{algorizmus} beweist, dass das Bewusstsein, dass \emph{Algorizmus} der Name eines Mannes sei, bei dem Verfasser der Abhandlung schon verloren gegangen war. Er hielt offenbar dieses Wort f{\"u}r den Namen der Rechenkunst selbst. \smallskip \end{quote}

@@

Algorithms should not only---which is obvious---be correct, that is, give a true output, but also, as Donald Knuth writes \parencite*[p.7]{Knuth1997}: \begin{quote} \small we want good algorithms in some loosely defined aesthetic sense. One criterion [\dots] is the length of time taken to perform the algorithm [\dots] Other criteria are adaptability of the algorithm to computers, its simplicity and elegance, etc. \end{quote}

Gregory Chaitin \parencite*[p.27] {Chaitin2005} specifies the concept of elegance in the program: \begin{quote} \small a program is `elegant', by which I mean that it's the smallest possible program for producing the output that it does. \end{quote} At the same time, he adds that: \begin{quote} \small I'll show you can't prove that a program is `elegant'---such a proof would solve the Halting problem. \end{quote}

@@

Let us add that the word ``computer'' was still in the 19\textsuperscript{th} century, and even in 1936---when Turing published \emph{On Computable Numbers, with an Application to the {E}ntscheidungsproblem} \parencite*{Turing1936}---used to indicate an official who was doing cumbersome numerical calculations \parencite[p.446]{CopelandSprevakShagrir2017}. Thus understood, ``computer'' would denote a reckoner, in Polish: rachmistrz. Polish texts in which ``computer'' is translated into ``komputer'' are devoid of any associations present in English texts. In particular, the association of ``komputer'' with ``rachmistrz'' is important for the correct understanding of the  Turing texts.

\section{Reality and information} 
Information is the content of our knowledge.\footnote{Stacewicz \parencite*[\S 1. Różne wymiary informacji (Different dimensions of information)]{MarciszewskiStacewicz2011} excellently discusses the concept of information and its relationship with knowledge.}
According to Luciano Floridi \parencite*{Floridi2008}, the pioneer of the philosophy of information, reality in itself is not a source but a resource for knowledge.  Stehphen Wolfram \parencite*[p.389]{Wolfram2002} states: \begin{quote} \small [M]atter is merely our way of representing to ourselves things that are in fact some pattern of information, but we can also say that matter is the primary thing and that information is our representation of that. It makes little difference, I don’t think there’s a big distinction---if one is right that there’s an ultimate model for the representation of universe in terms of computation. \end{quote} The retrieved information must be represented somehow. Representation enables it to be stored, communicated and processed. Each piece of information may be zero-one encoded. The way of representation is subordinated to the purpose and what it is supposed to serve. As John Wheeler \parencite*{Wheeler1989} puts it: \begin{quote} \small every physical quantity, every it, derives its ultimate significance from bits, binary yes-or-no indications. \end{quote} This idea can be summed up in short: it from bit, where ``it'' is what exists and ``bit'' refers to information.

Konrad Zuse \parencite*[p.5]{Zuse2012a} developing the concept of digitized spatial relations, the idea of understanding the universe as a computer, assigns an important role to the concept of information: \begin{quote} \small In current expanded usage, the term ``compute'' is identical with ``information processing.'' By analogy, the terms ``computer'' and ``information-processing machine'' may be taken as identical. \end{quote} Zuse was the first to suggest that the physical states of the universe are computed by the universe itself. He pointed to cellular automata. The concept of cellular automata was developed by John von Neumann in connection with his search for similarities between computers and the central nervous system \parencite{vonNeumann1951,vonNeumann1958,vonNeumann2012, Neumann1966,Shannon1958}.

@@

If thinking is a calculation, and the world is made of numbers, we will arrive at all the truth that we can arrive at by calculating. Thus \parencite[p.200]{Leibniz1890}\footnote{Similar statements can be found in other texts of the cited volume, for example on pages: 26, 64-65, 125.}: \begin{quote} \small \emph{Quo facto, quando orientur controversiae, non magis disputatione opus erit inter duos philosophos, quam inter duos Computistas. Sufficiet enim calamos in manus sumere sedereque ad abacos, et sibi mutuo (accito si placet amico) dicere: \ c a l c u l e m u s.}

\smallskip \end{quote}

The ontological thesis about the world as created by $1$ with $0$ has opened up new perspectives for combining the concept of information with metaphysics. In praising his binary arithmetic, Leibniz \parencite*{Leibniz1990} said: \begin{quote} \small \emph{tamen ubi Arithmeticam meam Binariam excogitavi, antequam Fohianorum characterum in mentem venirent, pulcherrimam in ea latereis judicavi ex nuhinem origin reisavi potentiam summae Unitatis, seu Dei.}

\end{quote} This idea fascinated Leibniz so much that he passed it on to Father Grimaldi, a mathematician at the court of the Emperor of China, in the hope that with it he would convert the Emperor and with him to christianize of all China \parencite{list1697}.

Calculating is an activity in which a machine can replace a human being. In 1685, when discussing the value for astronomers of a calculating machine he invented in 1673, more efficient than Pascal and performing all basic arithmetic operations, Leibniz \parencite*[p.181]{Leibniz1685} wrote that \parencite[Ch. I: Leibniz's Dream]{Davis2001}: \begin{quote} \small For it is unworthy of excellent men to lose hours like slaves in the labor of calculation which could safely be relegated to anyone else if the machine were used. \end{quote}

Charles Babbage, when he and his friend were preparing math tables, noticing a lot of errors, was frustrated and shouted \parencite{Swade2002}: \begin{quote} \small I wish to God these calculations had been executed by steam! \end{quote} Konrad Zuse in an interview with Uta Merzbach in 1978 said that when he had to do tedious engineering calculations, the think\footnote{Konrad Zuse interviewed by Uta Merzbach in 1968 (Computer Oral History Collection, Archives Center, National Museum of American History, Washington DC).}: \begin{quote} \small It's beneath a man. That should be accomplished with machines. \end{quote} motivated him to understand the work of building a computer \parencite[p.449]{CopelandSprevakShagrir2017}.

@@

In the case of the Galileo paradigm, the natural question is who is calculating. In the case of the Turing paradigm, the answer to the question of what processes information that makes up the world is simple: the world. An algorithm is part of the world just as data and programs are part of a computer, and just as data and programs are both encoded. Let us repeat the sentence which expresses this: \begin{quote} \small \emph{cum mundus calculat et algorithmum exercet, fit mundus.} \end{quote}

>>>>

\begin{artengenv}{Kazimierz Trzęsicki}
	{Perspective on Turing Paradigm}
	{Perspective on Turing Paradigm}
	{Perspective on Turing Paradigm}
	{University of Białystok}
	{Scientific knowledge is acquired according to some paradigm. Galileo wrote that the ``book of nature'' was written in mathematical language and could not be understood unless one first understood the language and recognized the characters with which it was written. It is argued that Turing planted the seeds of a new paradigm. According to the Turing Paradigm, the ``book of nature'' is written in algorithmic language, and science aims to learn how the algorithms change the physical, social, and human universe. Some sources of the Turing Paradigm are pointed out, and a few examples of the application of the Turing Paradigm are discussed.
	}
	{Galileo Galilei, Alan Turing, Konrad Zuse, zero, Arabic numeral, paradigm, mathematics, algorithmics.}

\epigraph{{\footnotesize \textit{Tolle numerum omnibus rebus, et omnia pereunt.} \normalfont{[Take from all things their number and all shall perish.]}}}{
{\footnotesize
%St. Isidore of Seville, Patron saint for the Internet, \emph{Etymologiarum sive originum}
\begin{flushleft}
\parencite[Liber III, \textit{De mathematica}, IV. \textit{Quid praestent numeri}]{Izydor1911}
\end{flushleft}
}\hfill \phantom{}}

@@

\lettrine[loversize=0.13,lines=2,lraise=-0.03,nindent=0em,findent=0.2pt]%
{S}{}cience is developed and created  according to a pattern, a paradigm. Paradigm is historically mutable. The place of one paradigm is taken by the one which enables a fuller understanding and a better description of the information obtained. The Aristotle paradigm of natural knowledge has been replaced by the paradigm  we used to tie with Galileo. According to this paradigm modern science was created. Appointed by Aristotele paradigm of logic lasted until Gottlob Frege. The Greek paradigm of mathematics has been replaced by a paradigm that we can associate with Descartes.

@@

It is usually stated that binary notation was invented and first formally proposed by Leibniz as an illustration of his dualistic philosophy, but already around 1600 the notation was used by  an English astronomer Thomas Harriot. John Shirley \parencite*{Shirley1951} writes about his achievements:
\myquote{
the mathematical papers of Thomas Harriot (1560--1621) show clearly that Harriot not only experimented with number systems, but also understood clearly the theory and practice of binary numeration nearly a century before Leibniz's time.
} 
Several manuscripts of the legacy of Thomas Harriot are evidence that he is probably the first inventor of binary system. He uses $0$ and $1$ and shows examples how to convert expressions  written in the decimal system to expressions written in the binary system and \emph{vice versa}. He demonstrates the basic arithmetic operations, too \parencite{Ineichen2008}. As the first text on the binary system Ineichen  points the two-volume work \emph{Mathesis biceps vetus et nova} \parencite*{Lobkowitz1670} by Juan Caramuel y Lobkowitz (Ioannis Caramuelis). In connection with these works by Harriot and Caramuel, the question is raised as to whether Leibniz plagiarized. This question is answered in the affirmative \parencite{Ares2018}.

@@

The use of the binary system in computers was finally determined by \emph{Burks-Goldstine-Von Neuman Report} to U.S. Army Ordnance Department (finished June 28, 1946) reprinted in \parencite*[p.105]{BurksGoldstineNeumann1947}, in which we read:
\myquote{
An additional point that deserves emphasis is this: An important part of the machine is not arithmetical, but logical in nature. Now logics, being a yes-no system, is fundamentally binary. Therefore, a binary arrangement of the arithmetical organs contributes very significantly towards a more homogeneous machine, which can be better integrated and is more efficient.\footnote{NB. This explanatory passage was not present in first edition of the report \parencite[cf.][p.13]{burks_preliminary_1946}, but was added in later editions (editor's footnote).}
}

@@

The idea that everything is made of $0$ and $1$ is the reason why one of the creators of algorithmic information theory, Gregory Chaitin---as he writes not quite seriously---proposes to name the basic unit of information not ``bit'' but ``leibniz'' \parencites{Chaitin2004}[cf.][]{Trzesicki2006a}:
\myquote{
all of information theory derives from Leibniz, for he was the first to emphasize the creative combinatorial potential of the $0$ and $1$ bit, and how everything can be built up from this one elemental choice, from these two elemental possibilities. So, perhaps not entirely seriously, I should propose changing the name of the unit of information from the bit to the leibniz!
}
The unit ``leibniz'' could be the unit (parcel) that Hobbes \parencite*[Chapter V. Of Reason, and Science]{Hobbes1651} wrote about:
\myquote{
When a man reasoneth, hee does nothing els but conceive a summe totall, from Addition of parcels.
}

Leibniz was convinced that the world was organized according to the rules of mathematics. This thought is summarized in the sentence \parencite*[p.191] {Leibniz1677}\footnote{More on this entry in the margin of the essay \emph{Dialogus} \parencite{Leibniz1677} see \parencite{Kopania2018}.}:
\myquote{
\emph{Cum Deus calculat et cogitationem exercet, fit mundus}.
}
Mathematics is a tool of the World Constructor, and numbers are the material the world is made of.

@@

The shaping of the modern paradigm of science in what was then called ``natural philosophy'' was in fact a revival of the concept of Archimedes \parencite[pp.71, 77] {Heller2013}. This idea continued in the Middle Ages. For Roger Bacon (1214--1292) there are four great sciences without which others cannot be known and the meaning of things cannot be understood. And when they are known, then wisdom will be attained without difficulty and labor, not only in the teachings of man, but also in the divine ones. And the possibilities of each of these sciences  are revealed not only because of the wisdom itself, but also in relation to the above-mentioned one. In  \emph{Opus Majus} \parencite*[Pars Quarta, Distinctio Prima, Capitulum I]{Bacon2010a} Roger Bacon emphasizes that:
\myquote{
Of these sciences the gate and key is mathematics, which the saints discovered at the beginning of the world, as I shall show, and which has always been used by all the saints and sages more than all other sciences. Neglect of mathematics works injury to all knowledge, since he who is ignorant of it cannot know the other sciences or the things of this world. And what is worse, men who are thus ignorant are unable to perceive their own ignorance and so do not seek a remedy.
}
About the place and role of experiments in \emph{De scientia experimentorum: que dicitur dignior Omnibus Partibus Philosophie Naturalis de Perspective: Et ideo notanda est maxime}, a part of \emph{Opus Tertium} \parencite*{Bacon1912}, he wrote that the strongest argument proves nothing so long as the conclusions are not verified by experience. Experimental science is the queen of sciences, and the goal of all speculation.

Galileo justifies heliocentrism by referring to the exegesis of \emph {Bible} based on the doctrine of St. Augustine, in particular his \emph{De Genesi ad litteram}
%PP \footnote{\url{http://www.inters.org/galilei-madame-christina-Lorraine} [02.04.2019]} 
\parencites{galileo_galilei_letter_1615}[cf.][p.73]{Sibley2013}. In this tradition, which Galileo finds explicitly, the book of nature should be read with mathematical tools rather than those of scholastic philosophy. The book of nature was written in the language of mathematics, and therefore must be interpreted by mathematicians, not theologians. The book of nature as a mathematics contains truths that cannot be discussed.

\pagebreak
Galileo \parencite*[p.4]{Galileo1960} writes: \myquote{Philosophy is written in this grand book, the universe, which stands continually open to our gaze. But the book cannot be understood unless one first learns to comprehend the language and read the letters in which it is composed. It is written in the language of mathematics, and its characters are triangles, circles, and other geometric figures without which it is humanly impossible to understand a single word of it; without these, one wanders
about in a dark labyrinth.} Galileo recommends learning the language of mathematics because it is the language spoken by God \parencite{Wouk2010,Strogatz2019}.

Hall \parencite*[p.97]{Hall1956}  maintains that: \myquote{ Galileo's greatest fame is as an astronomer, yet in intellectual quality and weight his one treatise on mechanics almost outweighs all the rest of his writings. Although his book on cosmology became notorious, and had a more general public influence, it had no comparable effect upon the future development of scientific astronomy, for its polemics were suited only to its own age. The contradiction here is more apparent than genuine. Though formally divided between two branches, Galileo's creative activity in science was a unity, not twofold: it was a unity in laying down revised principles of procedure in science, and again in its specific exemplification of these principles, since Galileo saw that the science of motion and the just appraisal of the results of observational astronomy were the twin keys to an understanding of the universe.}

@@

Newton creates calculus because it is the language with which the book of nature is written. Also Leibniz creates calculus. As an aside, let us add that Newton accused him of plagiarism \parencite{Sonar2018}. It just so happens that Leibniz, the genius of creating symbols \parencite['The Symbol Master', cf.][pp.165--168]{Mazur2014}---having a greater understanding of the choice of language---gave his version a linguistic representation that resulted in the development of which failed with the approach proposed by Newton. Charles Babbage, the creator of the first (mechanical) programmable computer, noticing the delay of English mathematics in relation to French mathematics, undertook to translate French texts from mathematics \parencite{Trzesicki2006c}. He wrote \parencite*{Babbage1864,Babbage2008}: \myquote{Under these circumstances it was not surprising that I should perceive and be penetrated with the superior power of the notation of Leibniz. }

@@

Later the idea of God (God of Spinoza) as a ``mathematician'' is proclaimed by Einstein \parencite[p.279]{Infeld1980}: \myquote{God does not care about our mathematical difficulties. He integrates empirically.} This is according to Heller \parencite*[p.41]{Heller2014}: \myquote{A fundamental hypothesis, tacitly taken in the very method of modern mathematized empirical science [which] states that there is nothing in the material world that cannot be mathematically described.}

@@

Robert of Chester, who was the first translator into Latin of the now-lost book al-Khwarizmi \parencite[p.411]{Menninger1969}, his translation---found in the 19\textsuperscript{th} century---begins with the words: \myquote{\emph {Dixit Algoritmi: laudes deo rectori nostro atque defensori dicmus dignas.}} Around 1143 \parencite[p.411]{Menninger1958} an abstract was made of this work today known as \emph{Salem Codex}  \parencite{Cantor1865}. At the beginning we read: \myquote{\emph {Incipit liber algorizmi: omnis sapientia sive scientia a domine Deo; sicut scriptum est: Hoc quod continent omnia scientiam habet, et iterum: Omnia in mensura et pondere et numero constituisti.}} The grammatical form used proves that the author was not aware that it was a surname \parencite[p.14, footnote 1]{Cantor1865}. In this text, the name ``algorizmi'' was for the first time---in the preserved literature---used to mark a procedure: \myquote{Der Gebrauch des Nominativus \emph{algorizmus} beweist, dass das Bewusstsein, dass \emph{Algorizmus} der Name eines Mannes sei, bei dem Verfasser der Abhandlung schon verloren gegangen war. Er hielt offenbar dieses Wort f{\"u}r den Namen der Rechenkunst selbst. \smallskip}

@@

Algorithms should not only---which is obvious---be correct, that is, give a true output, but also, as Donald Knuth writes \parencite*[p.7]{Knuth1997}: \myquote{we want good algorithms in some loosely defined aesthetic sense. One criterion [\ldots] is the length of time taken to perform the algorithm [\ldots] Other criteria are adaptability of the algorithm to computers, its simplicity and elegance, etc.}

Gregory Chaitin \parencite*[p.27] {Chaitin2005} specifies the concept of elegance in the program: \myquote{a program is `elegant', by which I mean that it's the smallest possible program for producing the output that it does.} At the same time, he adds that: \myquote{I'll show you can't prove that a program is `elegant'---such a proof would solve the Halting problem.}

@@

Let us add that the word ``computer'' was still in the 19\textsuperscript{th} century, and even in 1936---when Turing published \emph{On Computable Numbers, with an Application to the {E}ntscheidungsproblem} \parencite*{Turing1936}---used to indicate an official who was doing cumbersome numerical calculations \parencite[p.446]{CopelandSprevakShagrir2017}. Thus understood, ``computer'' would denote a reckoner, in Polish: rachmistrz. Polish texts in which ``computer'' is translated into ``komputer'' are devoid of any associations present in English texts. In particular, the association of ``komputer'' with ``rachmistrz'' is important for the correct understanding of the  Turing texts.
\enlargethispage{1.5\baselineskip}

\section{Reality and information} 
Information is the content of our knowledge.\footnote{Stacewicz \parencite*[\S 1]{MarciszewskiStacewicz2011} excellently discusses the concept of information and its relationship with knowledge.}
According to Luciano Floridi \parencite*{Floridi2008}, the pioneer of the philosophy of information, reality in itself is not a source but a resource for knowledge.  Stehphen Wolfram \parencite*[p.389]{Wolfram2002} states:

\myquote{[M]atter is merely our way of representing to ourselves things that are in fact some pattern of information, but we can also say that matter is the primary thing and that information is our representation of that. It makes little difference, I don’t think there’s a big distinction---if one is right that there’s an ultimate model for the representation of universe in terms of computation.} The retrieved information must be represented somehow. Representation enables it to be stored, communicated and processed. Each piece of information may be zero-one encoded. The way of representation is subordinated to the purpose and what it is supposed to serve. As John Wheeler \parencite*{Wheeler1989} puts it: \myquote{every physical quantity, every it, derives its ultimate significance from bits, binary yes-or-no indications.} This idea can be summed up in short: it from bit, where ``it'' is what exists and ``bit'' refers to information.

Konrad Zuse \parencite*[p.5]{Zuse2012a} developing the concept of digitized spatial relations, the idea of understanding the universe as a computer, assigns an important role to the concept of information: \myquote{In current expanded usage, the term ``compute'' is identical with ``information processing.'' By analogy, the terms ``computer'' and ``information-processing machine'' may be taken as identical.} Zuse was the first to suggest that the physical states of the universe are computed by the universe itself. He pointed to cellular automata. The concept of cellular automata was developed by John von Neumann in connection with his search for similarities between computers and the central nervous system \parencite{vonNeumann1951,vonNeumann1958,vonNeumann2012, Neumann1966,Shannon1958}.

@@

If thinking is a calculation, and the world is made of numbers, we will arrive at all the truth that we can arrive at by calculating. Thus \parencite[p.200]{Leibniz1890}\footnote{Similar statements can be found in other texts of the cited volume, for example on pages: 26, 64-65, 125.}: \myquote{\emph{Quo facto, quando orientur controversiae, non magis disputatione opus erit inter duos philosophos, quam inter duos Computistas. Sufficiet enim calamos in manus sumere sedereque ad abacos, et sibi mutuo (accito si placet amico) dicere: \ c a l c u l e m u s.}

\smallskip}

The ontological thesis about the world as created by $1$ with $0$ has opened up new perspectives for combining the concept of information with metaphysics. In praising his binary arithmetic, Leibniz \parencite*{Leibniz1990} said: \myquote{\emph{tamen ubi Arithmeticam meam Binariam excogitavi, antequam Fohianorum characterum in mentem venirent, pulcherrimam in ea latereis judicavi ex nuhinem origin reisavi potentiam summae Unitatis, seu Dei.}
}
This idea fascinated Leibniz so much that he passed it on to Father Grimaldi, a mathematician at the court of the Emperor of China, in the hope that with it he would convert the Emperor and with him to christianize of all China \parencite{list1697}.

Calculating is an activity in which a machine can replace a human being. In 1685, when discussing the value for astronomers of a calculating machine he invented in 1673, more efficient than Pascal and performing all basic arithmetic operations, Leibniz \parencite*[p.181]{Leibniz1685} wrote that \parencite[Ch. I: Leibniz's Dream]{Davis2001}: \myquote{For it is unworthy of excellent men to lose hours like slaves in the labor of calculation which could safely be relegated to anyone else if the machine were used.}

Charles Babbage, when he and his friend were preparing math tables, noticing a lot of errors, was frustrated and shouted \parencite{Swade2002}: \myquote{I wish to God these calculations had been executed by steam!} Konrad Zuse in an interview with Uta Merzbach in 1978 said that when he had to do tedious engineering calculations, the think\footnote{Konrad Zuse interviewed by Uta Merzbach in 1968 (Computer Oral History Collection, Archives Center, National Museum of American History, Washington DC).}: \myquote{It's beneath a man. That should be accomplished with machines.} motivated him to understand the work of building a computer \parencite[p.449]{CopelandSprevakShagrir2017}.

@@

In the case of the Galileo paradigm, the natural question is who is calculating. In the case of the Turing paradigm, the answer to the question of what processes information that makes up the world is simple: the world. An algorithm is part of the world just as data and programs are part of a computer, and just as data and programs are both encoded. Let us repeat the sentence which expresses this: \myquote{\emph{cum mundus calculat et algorithmum exercet, fit mundus.}}
